---
title: "Lab 3 - Data tidying and joining"
author: "Luke Owen"
format:
  html: default
  pdf: default
  
categories: "Lab"
---


## Packages

In this lab we will work with the **tidyverse** package, which is a collection of packages for doing data analysis in a "tidy" way.

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1

**Inflation across the world**

For this part of the analysis you will work with inflation data from various countries in the world over the last 30 years.

```{r}
#| label: load-country-data
#| message: false

country_inflation <- read_csv("data/country-inflation.csv")
write_csv(country_inflation, "c.inflation.csv")



```

### Question 1

Get to know the data.

a.  `glimpse()` at the `country_inflation` data frame and answer the following questions based on the output. How many rows does `country_inflation` have and what does each row represent? How many columns does `country_inflation` have and what does each column represent?

```{r}
glimpse(country_inflation)
```

This dataset has 44 rows, and 30 columns. Each row represents one country and its inflation rate for each year from 1993–2021. While the columns represent yearly inflation, except for column 1 which is held for the country name.

b.  Display a list of the countries included in the dataset.

```{r}
country_inflation |>
  pull(country)
```

### Question 2

Which countries had the top three highest inflation rates in 2021? Your output should be a data frame with two columns, country and 2021, with inflation rates in descending order, and three rows for the top three countries. Briefly comment on how the inflation rates for these countries compare to the inflation rate for United States in that year.

```{r}
inflation_2021 <- country_inflation |>
  select(country, `2021`) |>
  arrange(desc(`2021`)) |>
  slice_head(n = 3)

inflation_2021
```

The countries with the three highest inflation rates in 2021 were Argentina (48%), Turkey (19%), and Nigeria (17%).   In comparison, the United States had a much lower inflation rate of about 4.7% in 2021, meaning inflation in these countries was several times higher than in the United States that year.


### Question 3

In a single pipeline,

-   calculate the ratio of the inflation in 2021 and inflation in 1993 for each country and store this information in a new column called `inf_ratio`,

```{r}
country_inflation |>
  mutate(inf_ratio = `2021` / `1993`) |>
  arrange(desc(inf_ratio)) |>
  select(country, inf_ratio)
```

The country with the largest inflation change between 1993 and 2021 is New Zealand, as it has the highest inflation ratio. Since the ratio is greater than 1, this indicates that inflation increased between 1993 and 2021 in this country.

### Question 4

Reshape (pivot) country_inflation such that each row represents a country/year combination, with columns country, year, and annual_inflation. Then, display the resulting data frame and state how many rows and columns it has.

```{r}
inflation_long <- country_inflation |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "annual_inflation",
    names_transform = as.numeric
  )

inflation_long
```

The dataset was reshaped from wide format to long format using pivot_longer(), where each row represents a unique country and year combination. The resulting data frame contains the variables country, year, and annual_inflation, and the year variable was converted to numeric using names_transform = as.numeric. The resulting data frame contains 1276 rows and 3 columns.


### Question 5

Use a separate, single pipeline to answer each of the following questions.

Requirement: Your code must use the `filter()` function for each part, not `arrange()`.

a.  What is the highest inflation rate observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(annual_inflation == max(annual_inflation, na.rm = TRUE))
```
Brazil had the highest inflation observed in 1994, at 2075.9%. 

b.  What is the lowest inflation rate observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(annual_inflation == min(annual_inflation, na.rm = TRUE))
```

Irleand had the lowest inflation observed in 2009, at -4.5%. 

c.  Putting (a) and (b) together: What are the highest and the lowest inflation rates observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(
    annual_inflation == max(annual_inflation, na.rm = TRUE) |
    annual_inflation == min(annual_inflation, na.rm = TRUE)
  )
```


### Question 6

a\. Create a vector called `countries_of_interest` which contains the names of up tp five countries you want to visualize the inflation rates for over the years. 


```{r}
countries_of_interest <- c(
  "United States",
  "Canada",
  "United Kingdom",
  "Japan",
  "Belgium"
)
```

I selected these countries because they represent a mix of large developed economies and allow for comparison of inflation trends across different regions of the world. (Lastly, Belgium was chosen to have a better graph to present).

b\. In a single pipeline, filter your reshaped dataset to include only the `countries_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` countries in the data frame you created.

```{r}
inflation_selected <- inflation_long |>
  filter(country %in% countries_of_interest)
```

```{r}
  inflation_selected |>
  distinct(country)
```


### Question 7

Using your data frame from the previous question, create a plot of annual inflation vs. year for these countries. Then, in a few sentences, describe the patterns you observe in the plot, particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of these countries economies.

```{r}
ggplot(
  inflation_selected,
  aes(
    x = year,
    y = annual_inflation,
    color = country,
    shape = country,
    group = country
  )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Annual inflation rate",
    color = "Country",
    shape = "Country",
    title = "Inflation trends from 1993 to 2021"
  ) +
  theme_minimal()
```

The plot shows that inflation rates for the selected countries are generally fairly low and stable over time, although there are occasional spikes in certain years. This is not especially surprising, since many large developed economies tend to maintain relatively moderate inflation under normal economic conditions. 

## Part 2

**Inflation in the US**

The OECD defines inflation as follows:

> Inflation is a rise in the general level of prices of goods and services that households acquire for the purpose of consumption in an economy over a period of time.
>
> The main measure of inflation is the annual inflation rate which is the movement of the Consumer Price Index (CPI) from one month/period to the same month/period of the previous year expressed as percentage over time.
>
> Source: [OECD CPI FAQ](https://www.oecd.org/sdd/prices-ppp/consumerpriceindices-frequentlyaskedquestionsfaqs.htm#1)

CPI is broken down into 12 divisions such as food, housing, health, etc. Your goal in this part is to create another time series plot of annual inflation, this time for US only.

The data you will need to create this visualization is spread across two files:

-   `us-inflation.csv`: Annual inflation rate for the US for 12 CPI divisions. Each division is identified by an ID number.
-   `cpi-divisions.csv`: A "lookup table" of CPI division ID numbers and their descriptions.

Let's load both of these files.

```{r}
#| label: load-us-data
#| message: false

us_inflation <- read_csv("data/us-inflation.csv")
cpi_divisions <- read_csv("data/cpi-divisions.csv")
```

### Question 8

a\. How many columns and how many rows does the `us_inflation` dataset have? What are the variables in it? Add a brief (1-2 sentences) narrative summarizing this information.

b\. How many columns and how many rows does the `cpi_divisions` dataset have? What are the variables in it? Add a brief (1-2 sentences) narrative summarizing this information.

c\. Create a new dataset by joining the `us_inflation` dataset with the `cpi_division_id` dataset.

-   Determine which type of join is the most appropriate one and use that.

-   Note that the two datasets don't have a common variable. Review the help for the join functions to determine how to use the `by` argument when the names of the variables that the datasets should be joined by are different.

-   Use a short but informative name for the joined dataset, and do not overwrite either of the datasets that go into creating it.

Then, find the number of rows and columns of the resulting dataset and report the names of its columns. Add a brief (1-2 sentences) narrative summarizing this information.

### Question 9

a\. Create a vector called `divisions_of_interest` which contains the descriptions or IDs of CPI divisions you want to visualize. Your `divisions_of_interest` should consist of no more than five divisions. If you're using descriptions, make sure that the spelling of your divisions matches how they appear in the dataset. Then, in 1-2 sentences, state why you chose these divisions.

::: callout-tip
Refer back to the guidance provided in [Question 6] if you're not sure how to create this vector.
:::

b\. In a single pipeline, filter your reshaped dataset to include only the `divisions_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` divisions in the data frame you created.

### Question 10

Using your data frame from the previous question, create a plot of annual inflation vs. year for these divisions. Then, in a few sentences, describe the patterns you observe in the plot, particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of inflation rates in the US over the last decade.

-   Data should be represented with points as well as lines connecting the points for each division.
-   Each division should be represented by a different color line and different color and shape points.
-   Axes and legend should be properly labeled.
-   The plot should have an appropriate title (and optionally a subtitle).
-   Plot should be customized in at least one way – you could use a different than default color scale, or different than default theme, or some other customization.
-   If your legend has labels that are too long, you can try moving the legend to the bottom and stack the labels vertically. *Hint:* The `legend.position` and `legend.direction` arguments of the `theme()` functions will be useful.

```{r}
ggplot() +
  theme(
    legend.position = "bottom", 
    legend.direction = "vertical"
  )
```

::: render-commit-push
If you haven't yet done so since Part 1, now is a good time to render, commit, and push. Make sure that you commit and push all changed documents and your Git pane is completely empty before proceeding.
:::

# Wrap-up

## Submission

Once you are finished with the lab, you will submit your final PDF document to Canvas.

::: callout-warning
Before you wrap up the assignment, make sure all of your documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

You must turn in a PDF file to the Canvas page by the submission deadline to be considered "on time".
:::

To submit your assignment:

-   Go to <http://www.Canvas.com> and click *Log in* in the top right corner.
-   Click *School Credentials* $\rightarrow$ *Duke NetID* and log in using your NetID credentials.
-   Click on your *REN R 213* course.
-   Click on the assignment, and you'll be prompted to submit it.
-   Mark all the pages associated with question. All the pages of your lab should be associated with at least one question (i.e., should be "checked").

::: callout-important
## Checklist

Make sure you have:

-   attempted all questions
-   rendered your Quarto document
-   committed and pushed everything to your GitHub repository such that the Git pane in RStudio is empty
-   uploaded your PDF to Canvas
-   selected pages associated with each question on Canvas
:::

## Grading

The lab is graded out of a total of 36 points.

You can earn up to 4 points on each question:

-   4: Response shows excellent understanding and addresses all or almost all of the rubric items.

-   3: Response shows good understanding and addresses most of the rubric items.

-   2: Response shows understanding and addresses a majority of the rubric items.

-   1: Response shows effort and misses many of the rubric items.

-   0: Response does not show sufficient effort or understanding and/or is largely incomplete.
