---
title: "Lab 3 - Data tidying and joining"
author: "Luke Owen"
format:
  pdf: default
  html: default
  
categories: "Lab"
---


## Packages

In this lab we will work with the **tidyverse** package, which is a collection of packages for doing data analysis in a "tidy" way.

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1

**Inflation across the world**

For this part of the analysis you will work with inflation data from various countries in the world over the last 30 years.

```{r}
#| label: load-country-data
#| message: false

country_inflation <- read_csv("data/country-inflation.csv")
write_csv(country_inflation, "c.inflation.csv")



```

### Question 1

Get to know the data.

a.  `glimpse()` at the `country_inflation` data frame and answer the following questions based on the output. How many rows does `country_inflation` have and what does each row represent? How many columns does `country_inflation` have and what does each column represent?

```{r}
glimpse(country_inflation)
```

This dataset has 44 rows, and 30 columns. Each row represents one country and its inflation rate for each year from 1993–2021. While the columns represent yearly inflation, except for column 1 which is held for the country name.

b.  Display a list of the countries included in the dataset.

```{r}
country_inflation |>
  pull(country)
```

### Question 2

Which countries had the top three highest inflation rates in 2021? Your output should be a data frame with two columns, country and 2021, with inflation rates in descending order, and three rows for the top three countries. Briefly comment on how the inflation rates for these countries compare to the inflation rate for United States in that year.

```{r}
inflation_2021 <- country_inflation |>
  select(country, `2021`) |>
  arrange(desc(`2021`)) |>
  slice_head(n = 3)

inflation_2021
```

The countries with the three highest inflation rates in 2021 were Argentina (48%), Turkey (19%), and Nigeria (17%).   In comparison, the United States had a much lower inflation rate of about 4.7% in 2021, meaning inflation in these countries was several times higher than in the United States that year.


### Question 3

In a single pipeline,

-   calculate the ratio of the inflation in 2021 and inflation in 1993 for each country and store this information in a new column called `inf_ratio`,

```{r}
country_inflation |>
  mutate(inf_ratio = `2021` / `1993`) |>
  arrange(desc(inf_ratio)) |>
  select(country, inf_ratio)
```

The country with the largest inflation change between 1993 and 2021 is New Zealand, as it has the highest inflation ratio. Since the ratio is greater than 1, this indicates that inflation increased between 1993 and 2021 in this country.

### Question 4

Reshape (pivot) country_inflation such that each row represents a country/year combination, with columns country, year, and annual_inflation. Then, display the resulting data frame and state how many rows and columns it has.

```{r}
inflation_long <- country_inflation |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "annual_inflation",
    names_transform = as.numeric
  )

inflation_long
```

The dataset was reshaped from wide format to long format using pivot_longer(), where each row represents a unique country and year combination. The resulting data frame contains the variables country, year, and annual_inflation, and the year variable was converted to numeric using names_transform = as.numeric. The resulting data frame contains 1276 rows and 3 columns.


### Question 5

Use a separate, single pipeline to answer each of the following questions.

Requirement: Your code must use the `filter()` function for each part, not `arrange()`.

a.  What is the highest inflation rate observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(annual_inflation == max(annual_inflation, na.rm = TRUE))
```
Brazil had the highest inflation observed in 1994, at 2075.9%. 

b.  What is the lowest inflation rate observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(annual_inflation == min(annual_inflation, na.rm = TRUE))
```

Irleand had the lowest inflation observed in 2009, at -4.5%. 

c.  Putting (a) and (b) together: What are the highest and the lowest inflation rates observed between 1993 and 2021? 

```{r}
inflation_long |>
  filter(
    annual_inflation == max(annual_inflation, na.rm = TRUE) |
    annual_inflation == min(annual_inflation, na.rm = TRUE)
  )
```


### Question 6

a\. Create a vector called `countries_of_interest` which contains the names of up tp five countries you want to visualize the inflation rates for over the years. 


```{r}
countries_of_interest <- c(
  "United States",
  "Canada",
  "United Kingdom",
  "Japan",
  "Belgium"
)
```

I selected these countries because they represent a mix of large developed economies and allow for comparison of inflation trends across different regions of the world. (Lastly, Belgium was chosen to have a better graph to present).

b\. In a single pipeline, filter your reshaped dataset to include only the `countries_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` countries in the data frame you created.

```{r}
inflation_selected <- inflation_long |>
  filter(country %in% countries_of_interest)
```

```{r}
  inflation_selected |>
  distinct(country)
```

### Question 7

Using your data frame from the previous question, create a plot of annual inflation vs. year for these countries. Then, in a few sentences, describe the patterns you observe in the plot, particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of these countries economies.

```{r}
ggplot(
  inflation_selected,
  aes(
    x = year,
    y = annual_inflation,
    color = country,
    shape = country,
    group = country
  )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Annual inflation rate",
    color = "Country",
    shape = "Country",
    title = "Inflation trends from 1993 to 2021"
  ) +
  theme_minimal()
```

The plot shows that inflation rates for the selected countries are generally fairly low and stable over time, although there are occasional spikes in certain years. This is not especially surprising, since many large developed economies tend to maintain relatively moderate inflation under normal economic conditions. 

## Part 2

**Inflation in the US**

The data you will need to create this visualization is spread across two files:

-   `us-inflation.csv`: Annual inflation rate for the US for 12 CPI divisions. Each division is identified by an ID number.
-   `cpi-divisions.csv`: A "lookup table" of CPI division ID numbers and their descriptions.

Let's load both of these files.

```{r}
#| label: load-us-data
#| message: false

us_inflation <- read_csv("data/us-inflation.csv")
write_csv(us_inflation, "us_inflation.csv")
cpi_divisions <- read_csv("data/cpi-divisions.csv")
write_csv(cpi_divisions, "cpi_divisions.csv")
```

### Question 8

a\. How many columns and how many rows does the `us_inflation` dataset have? What are the variables in it? Add a brief (1-2 sentences) narrative summarizing this information.

```{r}
glimpse(us_inflation)
```

The us_inflation dataset contains 132 rows and 4 columns. The variables in the dataset are country, year, annual_inflation,  cpi_division_id.

b\. How many columns and how many rows does the `cpi_divisions` dataset have? What are the variables in it? Add a brief (1-2 sentences) narrative summarizing this information.

```{r}
glimpse(cpi_divisions)
```

The cpi_division dataset contains 12 rows and 2 columns. The variables in the dataset are id and description.

c\. Create a new dataset by joining the `us_inflation` dataset with the `cpi_division_id` dataset.

```{r}
inflation_joined <- us_inflation |>
  left_join(
    cpi_divisions,
    by = c("cpi_division_id" = "id")
  )
```

Then, find the number of rows and columns of the resulting dataset and report the names of its columns. Add a brief (1-2 sentences) narrative summarizing this information.

```{r}
glimpse(inflation_joined)
```
The datasets were joined using a left join, which preserves all observations from the us_inflation dataset while adding the CPI division information from cpi_divisions. The resulting dataset contains 132 rows and 5 columns, and the variables are country, year, annual_inflation,cpi_division_id, and description.

### Question 9

a\. Create a vector called `divisions_of_interest` which contains the descriptions or IDs of CPI divisions you want to visualize. 

```{r}
divisions_of_interest <- c(
  "Food and non-Alcoholic beverages",
  "Housing, water, electricity, gas and other fuels",
  "Health",
  "Transport",
  "Education"
)
```

I chose these CPI divisions because they represent major categories of household spending and are likely to show meaningful differences in inflation over time. They also include categories that people commonly notice in everyday life, such as food, housing, and transportation.


b\. In a single pipeline, filter your reshaped dataset to include only the `divisions_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` divisions in the data frame you created.

```{r}
division_selected <- inflation_joined |>
  filter(description %in% divisions_of_interest)

division_selected |>
  distinct(description)
```

### Question 10

Using your data frame from the previous question, create a plot of annual inflation vs. year for these divisions. Then, in a few sentences, describe the patterns you observe in the plot, particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of inflation rates in the US over the last decade.

```{r}
ggplot(
  division_selected,
  aes(
    x = year,
    y = annual_inflation,
    color = description,
    shape = description,
    group = description
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    x = "Year",
    y = "Annual inflation rate",
    color = "CPI Division",
    shape = "CPI Division",
    title = "Annual inflation rates from 1993 to 2021"
  ) +
  theme_minimal()

```

